---
title: "Wrangling Data with Tidyr"
author: "Colleen McCamy"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

library(tidyverse)
library(here)
library(janitor)

```


## Read in WB data files

```{r}

wb_indicators <- read_csv(here::here("data", "wb_indicators.csv"), na = c("..", ""))

# ".." are missing but stored as something for not an empty cell this will give us the read in stage for the true NA, R will automatically see an empty cell as an NA

# reading in the metadata too

wb_metatdata <-read_csv(here("data", "wb_indicators_metadata.csv"))

```

# Tidying up the data

```{r}

# first going to gather all of these columns for the years using pivot longer
# 1. need to give it what columns do I want to gather together - via span of column names
# 2. next need to give it the column name for a new column
# 3. need to say what the value column name should be

wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = '2001 [YR2001]': '2020 [YR2020]', 
               names_to = "year",
               values_to = "indicator_value")

# year is classified as a character and we would want to address this class(wb_indicators_long$year)

```


### separate to clean up the year column

```{r}

# year is classified as a character and we would want to address this class(wb_indicators_long$year)

# want to take this original year column and split up what is in the column in year and year character

wb_clean <- wb_indicators_long |> 
  separate(col = year, into = c("year", "year_char"), sep = " ") |> 
  select(-year_char, -"Country Code", -"Series Code") |> 
  mutate(year = as.numeric((year))) |> 
  drop_na('Series Name') |> 
  pivot_wider(names_from = 'Series Name', values_from = indicator_value)

names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")

wb_clean_2 <- wb_clean |> 
  rename(my_year = year)

# first give column name
# then say where you want to separate the data (in to what column name)
# then say what you want to delineate from


# then piping into a function used to keep or include columns - filter is rows / select is column

# if r starts as a class then R won't update the column class

# using mutate to update an existing column or add a new column, to update an existing column include the same column title

# drop_na = any row where any value in a column in NA then the whole row will get dropped

# Pivot Wider wants two arguments = names from and values from

# renaming column names 

```

## Subsetting with filter

```{r}

wb_subset <- wb_clean |> 
  filter(country %in% c("Algeria", "Barbados", "Bulgaria", "Chile"))

# can check values in column country with unique()

ggplot(data = wb_subset, aes(x = year, y = co2_emissions_kt)) + geom_line(aes(color = country)) +
  facet_wrap(~country)

exclude_this <- c("Algeria", "Barbados", "Bulgaria", "Chile")

wb_example_grouping <- wb_clean |> 
  filter(!country %in% exclude_this)

# this is excluding items that you don't want to include

```

## Filtering Rows

Example 1: filtering to only keep exact matches

```{r}
# only filtering for height greater than 180

ex_1 <- starwars |> 
  filter(height > 180)

```

# exact match filtering

```{r}

ex_2 <- starwars |> 
  filter(eye_color == "blue")

ex_3 <- starwars |> 
  filter(homeworld == "Naboo")

```

Filtering to satisfy multiple conditions - for both to be true

```{r}

ex_4 <- starwars |> 
  filter(height > 180 & homeworld == "Tatooine")

ex_5 <- starwars |> 
  filter(species == "Human" & hair_color == "brown")


```

Filtering OR statements

```{r}

ex_6 <- starwars |> 
  filter(height > 180 | eye_color == "yellow")

ex_7 <- starwars |> 
  filter(homeworld == "Endor" | species == "Droid")
# keep rows if homeowrld is Endor or species is droid
```

Using the in operator

```{r}

ex_8 <- starwars |> 
  filter(eye_color == "blue" | eye_color == "red" | eye_color == "brown")

ex_9 <- starwars |> 
  filter(eye_color %in% c("blue", "brown", "red"))

# the in operator looks in each cell and ask is this cell blue, brown or red

# do not use the double equals as it is asking if the row equals vector item 1, while we want it for each cell to match blue, red or brown

ex_10 <- starwars |> 
    filter(eye_color == c("blue", "brown", "red"))
# NEVER DO THIS ^^^^^^^

ex_11 <- starwars |> 
  filter(homeworld %in% c("Tatooine", "Naboo", "Endor"), 
         height < 150)

```

Examples with the negate operator

```{r}

ex_12 <- starwars |> 
  filter(species != "Human")

ex_13 <- starwars |> 
  filter(!species %in% c("Human", "Ewok", "Wookiee"))

```

## `select()` The columns one = select or include columns

multiple ways to do this, indexing and to use the dollar sign but today we are using the select() function

```{r}

# Select by name 

ex_14 <- wb_clean |> 
  select(country, year, co2_emissions_kt)

# Select by range of column names

ex_15 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct)

# exculding thins from a range too

ex_16 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct, -access_electricity_pp)

```


```{r}

#can pipe through multiple sequences in all one items so you don't need to create intermediate stages
#always check at the different points of a pipe statement

ex_17 <- wb_clean |> 
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |> 
  select(country:fossil_fuel_cons_pct)

```

#Using select to rename and move things around

```{r}

ex_18 <- wb_clean |> 
  select(year_new = year,
         emissions = co2_emissions_kt,
         country_name = country)

# a bunch of things happened
# 1. kept the columns that you wanted to keep
# 2. renamed the columns,
# 3. re ordered the columns

```

```{r}

# select only columns for country, year and water stress and reorder as them as year, country water stress, then filter to only include observations for any three countries of you choosing

ex_19 <- wb_clean |> 
  select(year = year, country = country, water_stress = water_stress) |> 
  filter(country %in% c("Malawi", "Costa Rica", "Lebanon"))

# silent errors are way more scary

```

Renaming Columns

```{r}

ex_20 <- wb_clean |> 
  rename(ffcp = fossil_fuel_cons_pct, 
         ws = water_stress)

ex_21 <- wb_clean |> 
  rename(place = country, 
         electric = access_electricity_pp)

```

## `mutate()` update an exisiting column or add a new one

If you assign a NEW column name, it will add a new column to the end of your date frame. If you asign an existing column name, it will overwrite that column.

```{r}

ex_22 <- wb_clean |> 
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |> 
  relocate(co2_emissions_t, .after = co2_emissions_kt)

ex_23 <- wb_clean |> 
  mutate(yay = "YAY")

# can add a new column based on a specific value

ex_24 <- wb_clean |> 
  mutate(year = as.character(year))

ex_25 <- wb_clean |> 
  mutate(year = as.numeric(year))

# dplyr::relocate()

ex_26 <- wb_clean |> 
  relocate(water_stress, .after = year)



```


## `groub_by |> summarize` 
group_by : Creates groups under the hood - just using group by will do nothing
summarize : apply your functions to EACH GROUP and return them in a nice summary table

```{r}

# calculate average heights of 

ex_27 <- starwars |> 
  group_by(species, homeworld) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))

```


```{r}

#frequency table below - multiple observations that match across the way and counts them - we want to create a case table where each is it's own case

df <- tribble(
  ~ species, ~length, ~number,
  "lobsters", 12, 4,
  "lobsters", 14, 6,
  "lobsters", 15, 2,
)

df_case <- df |> 
  uncount(number)

```


## Count functions

```{r}

# count counts observations
# comnination of groub_by summarize n()

# count for different homeworld in starwars data set

starwars |> 
  count(homeworld, species)

# it is doing group by - summarize and n which n is the number of rows

starwars |> 
  group_by(homeworld, species) |> 
  summarize(size = n())

# you can do multiple calculations in summarize but you can only count the number of rows by groups in count

```

Practice example:

- start for wb_clean
- only keep observations for Nicaragua, Ecuador, Peru and Chile AND for years greater than 2005
- only keep columns for country, year and co2 emissions
- add a new column that is co2 emissions converted to megatons
- find the total co2 emissions in megatons for the years still in the data set by the country (with sum)


```{r}

practice_example <- wb_clean |> 
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile"), year > 2005) |> 
  select(country, year, co2_emissions_kt) |> 
  mutate(co2_emissions_mt = co2_emissions_kt / 1000) |> 
  group_by(country) |> 
  summarize(sum_co2_emissions_mt = sum(co2_emissions_mt, na.rm = TRUE))


  # can use min to check years less than
  
```

